---
title: "ST2195 Coursework — Part 1: Metropolis-Hastings MCMC"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 6.5, fig.height = 4.5)
set.seed(42)
```

# Target Distribution

The target distribution is:
$$f(x) = \frac{1}{2}e^{-|x|}, \quad x \in \mathbb{R}$$

This is a **Laplace (double exponential)** distribution with location 0 and scale 1.

- Theoretical mean: $0$
- Theoretical variance: $2$ &rarr; standard deviation $\sqrt{2} \approx 1.4142$

# Helper Functions

```{r functions}
# Log of target density: log f(x) = -|x| - log(2)
log_f <- function(x) {
  -abs(x) - log(2)
}

# Target density
f_target <- function(x) {
  0.5 * exp(-abs(x))
}

# Random Walk Metropolis-Hastings algorithm
# Returns vector x[0], x[1], ..., x[N]  (length N+1)
metropolis_hastings <- function(N, s, x0 = 0.0) {
  x    <- numeric(N + 1)
  x[1] <- x0
  for (i in seq(2, N + 1)) {
    x_star <- rnorm(1, mean = x[i - 1], sd = s)
    log_r  <- log_f(x_star) - log_f(x[i - 1])
    log_u  <- log(runif(1))
    # Accept/reject using log-space criterion (numerically stable)
    if (log_u < log_r) {
      x[i] <- x_star
    } else {
      x[i] <- x[i - 1]
    }
  }
  return(x)
}

cat("Helper functions defined.\n")
```

---

# Part (a): Simulation with N = 10 000, s = 1

```{r part_a_run}
library(ggplot2)

N  <- 10000L
s  <- 1.0
x0 <- 0.0

set.seed(42)
samples_full <- metropolis_hastings(N, s, x0)
samples      <- samples_full[-1]   # x1, ..., xN (exclude x0)

cat(sprintf("Generated %d samples (x1 to x%d).\n", length(samples), N))
```

## Histogram, KDE, and True f(x)

```{r part_a_plot, fig.width=6.5, fig.height=4.5, out.width="90%"}
x_range      <- seq(-8, 8, length.out = 1000)
true_df      <- data.frame(x = x_range, y = f_target(x_range))

# KDE via R's built-in density()
kde_res <- density(samples, from = -8, to = 8, n = 1000)
kde_df  <- data.frame(x = kde_res$x, y = kde_res$y)

samples_df <- data.frame(x = samples)

ggplot() +
  geom_histogram(
    data = samples_df, aes(x = x, y = after_stat(density)),
    bins = 60, fill = "steelblue", alpha = 0.40, color = "white"
  ) +
  geom_line(
    data = kde_df, aes(x = x, y = y, color = "KDE"),
    linewidth = 1.3
  ) +
  geom_line(
    data = true_df, aes(x = x, y = y, color = "True f(x)"),
    linewidth = 1.8
  ) +
  scale_color_manual(
    name   = "",
    values = c("KDE" = "dodgerblue", "True f(x)" = "red")
  ) +
  labs(
    title = sprintf("Random Walk Metropolis: N = %s, s = %.1f", format(N, big.mark = ","), s),
    x = "x", y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

## Monte Carlo Estimates

```{r part_a_stats}
sample_mean     <- mean(samples)
sample_sd       <- sd(samples)          # uses N-1 denominator (R default)
acceptance_rate <- mean(diff(samples_full) != 0)

cat("=======================================================\n")
cat("  Monte Carlo Estimates — Part 1(a)\n")
cat("=======================================================\n")
cat(sprintf("  Sample Mean:          %+.6f\n", sample_mean))
cat(sprintf("  Theoretical Mean:      0.000000\n"))
cat(sprintf("  Sample Std Dev:        %.6f\n", sample_sd))
cat(sprintf("  Theoretical Std Dev:   %.6f  [sqrt(2)]\n", sqrt(2)))
cat(sprintf("  Acceptance Rate:       %.4f\n", acceptance_rate))
cat("=======================================================\n")
```

---

# Part (b): Gelman-Rubin $\hat{R}$ Convergence Diagnostic

Given $J$ chains of length $N$ (excluding $x_0$):

$$M_j = \frac{1}{N}\sum_{i=1}^{N} x_i^{(j)}, \quad
V_j = \frac{1}{N}\sum_{i=1}^{N}\bigl(x_i^{(j)} - M_j\bigr)^2$$

$$W = \frac{1}{J}\sum_{j=1}^{J} V_j, \quad
M = \frac{1}{J}\sum_{j=1}^{J} M_j, \quad
B = \frac{1}{J}\sum_{j=1}^{J}(M_j - M)^2$$

$$\hat{R} = \sqrt{\frac{B + W}{W}}$$

Convergence is generally accepted when $\hat{R} < 1.05$.

## $\hat{R}$ Function

```{r rhat_function}
compute_rhat <- function(N, s, J, x0_values = NULL) {
  if (is.null(x0_values)) {
    x0_values <- runif(J, min = -5, max = 5)
  }

  # Generate J chains
  chains <- lapply(seq_len(J), function(j) {
    chain_full <- metropolis_hastings(N, s, x0_values[j])
    chain_full[-1]   # x1, ..., xN
  })

  # Chain means Mj
  Mj <- sapply(chains, mean)

  # Within-chain variances Vj  (N denominator as per coursework formula)
  Vj <- mapply(function(chain, mj) mean((chain - mj)^2), chains, Mj)

  # W: overall within-sample variance
  W <- mean(Vj)

  # M: overall mean  ;  B: between-chain variance
  M <- mean(Mj)
  B <- mean((Mj - M)^2)

  if (W < 1e-12) return(Inf)
  sqrt((B + W) / W)
}

cat("compute_rhat() defined (returns R-hat).\n")
```

## Fixed Parameters: N = 2000, s = 0.001, J = 4

```{r part_b_fixed}
N_b       <- 2000L
J         <- 4L
s_fixed   <- 0.001
x0_values <- c(-5.0, -2.0, 2.0, 5.0)   # spread initial values

set.seed(42)
rhat_fixed <- compute_rhat(N_b, s_fixed, J, x0_values)

cat(sprintf("R-hat for N=%d, s=%.3f, J=%d: %.4f\n", N_b, s_fixed, J, rhat_fixed))
if (rhat_fixed < 1.05) {
  cat("  -> Convergence achieved (R-hat < 1.05)\n")
} else {
  cat("  -> NOT converged (R-hat >= 1.05) — chains mixing poorly for this s\n")
}
```

## $\hat{R}$ over Grid of s Values

```{r part_b_plot, fig.width=6.5, fig.height=4.5, out.width="90%"}
set.seed(42)

s_grid      <- seq(0.001, 1.0, length.out = 50)
rhat_values <- sapply(s_grid, function(s) compute_rhat(N_b, s, J, x0_values))

grid_df     <- data.frame(s = s_grid, rhat = rhat_values)
converged_s <- grid_df$s[grid_df$rhat < 1.05]

plot_title    <- sprintf("R-hat over grid of s values  (N = %d, J = %d)", N_b, J)
plot_subtitle <- "Convergence threshold: R-hat < 1.05"

ggplot(grid_df, aes(x = s, y = rhat)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_point(color = "steelblue", size = 2.0) +
  geom_hline(
    yintercept = 1.05, color = "red", linetype = "dashed", linewidth = 1.3
  ) +
  geom_hline(
    yintercept = 1.00, color = "green4", linetype = "dotted", linewidth = 1.1
  ) +
  annotate("text", x = 0.70, y = 1.065,
           label = "hat(R) == 1.05", color = "red", size = 3.8, parse = TRUE) +
  {
    if (length(converged_s) > 0)
      geom_ribbon(
        data = subset(grid_df, rhat < 1.05),
        aes(ymin = 1, ymax = rhat), fill = "green4", alpha = 0.12
      )
  } +
  labs(
    title    = plot_title,
    subtitle = plot_subtitle,
    x        = "s  (proposal standard deviation)",
    y        = expression(hat(R))
  ) +
  theme_minimal(base_size = 12)
```

```{r part_b_summary}


cat("R-hat summary:\n")
cat(sprintf("  min R-hat = %.4f  at s = %.4f\n",
            min(rhat_values), s_grid[which.min(rhat_values)]))
cat(sprintf("  max R-hat = %.4f  at s = %.4f\n",
            max(rhat_values), s_grid[which.max(rhat_values)]))

if (length(converged_s) > 0) {
  cat(sprintf("  Converged s range: [%.4f, %.4f]\n",
              min(converged_s), max(converged_s)))
} else {
  cat("  No s values achieve convergence in the tested range.\n")
}
```
